---
title: "Classification Methods: Mixture Models"
subtitle: "Latent Class & Latent Profile Analysis"
format:
  revealjs:
    theme: simple
    css: clemson.css
    slide-number: true
    logo: clemson_coe.png
    transition: slide
    chalkboard: true
    preview-links: auto
execute:
  echo: false
  eval: true
---

## Lecture Outline

- Latent Class Analysis (LCA)
  - Underlying theory
  - Example analysis in R (`poLCA`)
  - Interpreting model parameters
  - Assessing model fit
- Latent Profile Analysis (LPA)
  - Underlying theory
  - Example analysis in R (`tidyLPA`)
  - Comparing class solutions

# Overview

## Clusters Versus Classes

- When a researcher mentions cluster analysis, they often mean one of the following:
  - K-means clustering
  - Hierarchical clustering using distance methods
  - Discriminant analysis
  - Taxometrics
- Much less often, latent class analysis is included in the group
  - Although it is also useful for detecting clusters of observations
- For this lecture, we will consider *clusters* and *classes* to be synonymous
  - Clustering is often used to *find* clusters whereas classification is used to classify observations into known clusters

## LCA Versus Other Methods

- Although we use the terms classes and clusters synonymously, LCA differs from other methods
- **LCA is a model-based method** for clustering
  - LCA fits a statistical model to the data to determine classes
- Other methods (K-means, hierarchical) do not explicitly state a statistical model
- Being model-based means we make **explicit, testable assumptions** about our data
  - Assumptions can be checked against the observed data

# Latent Class Analysis

## {.center}

::: {style="text-align: center; font-size: 2em;"}
Latent Class Analysis
:::

## LCA Introduction

- Latent class models are commonly attributed to Lazarsfeld and Henry (1968)
- The final number of classes is **not predetermined** prior to analysis
  - The number of classes is determined through comparison of fit statistics after fitting multiple models
  - The characteristics of each class are determined following the analysis
  - Similar to K-means and hierarchical clustering in this respect

## Variable Types Used in LCA

- As originally conceived, LCA uses:
  - A set of **binary-outcome variables** — values coded as 0 or 1. Examples include:
    - Test items — scored correct (1) or incorrect (0)
    - True/false questions
    - Presence/absence indicators
    - Any other binary outcome
- Extensions allow for **ordered categorical** or **nominal** indicators
  - This general form is still called LCA regardless of indicator type

## LCA Process

- For a specified number of classes, LCA estimates:
  1. **For each class:** the probability that each indicator equals one $\pi_{jc}$
  2. **For each observation:** the probability of belonging to each class $\alpha_{ic}$
     - These sum to 1 across classes for each person
     - This differs from K-means, where class membership is certain
  3. **Across observations:** the overall probability that any observation is in each class $\eta_c$

## LCA Estimation

- LCA estimation differs from other clustering methods:
  - Hierarchical clustering creates new distance matrices at each step
  - K-means shifts cases between clusters using distance metrics
  - **LCA uses distributional assumptions** to find classes — the distribution provides the measure of "distance"
- Estimation is performed via the **Expectation-Maximization (EM) algorithm**
  - E-step: compute posterior class probabilities
  - M-step: update model parameters to maximize log-likelihood
  - Iterate until convergence
- **Multiple random starting values** are recommended to avoid local maxima

## LCA Distributional Assumptions

- Because LCA uses binary-outcome variables, it relies on a **binary-outcome distribution**
- Within each latent class, the variables are assumed to:
  - Be **independent** of one another (*Local Independence*)
  - Follow a **Bernoulli distribution** marginally:

$$f(x_i) = \left(\pi_i \right)^{x_i} \left(1-\pi_i\right)^{(1-x_i)}$$

- The Bernoulli distribution describes a single binary event — like flipping a coin with probability $\pi$ of heads

## Bernoulli Distribution Illustration

- Consider a single binary test item, $X$:
  - Let $X = 1$ if a student answers correctly, $X = 0$ if incorrect
  - Suppose the probability of a correct response is $\pi = 0.75$

- If $X = 1$, the likelihood is:
$$f(x_i=1) = (0.75)^{1}(1-0.75)^{0} = 0.75$$

- If $X = 0$, the likelihood is:
$$f(x_i=0) = (0.75)^{0}(1-0.75)^{1} = 0.25$$

- For discrete-outcome variables, the likelihood equals the probability of the event occurring

## Independent Bernoulli Variables

- For **independent** binary variables, the joint probability is the product of the individual probabilities:

$$P(X_1=x_1, X_2=x_2,\ldots,X_J=x_J) = \prod_{j=1}^{J} \pi_j^{x_j} \left(1-\pi_j\right)^{\left(1-x_j\right)}$$

- In LCA, this independence assumption holds **within each class** (Local Independence)
  - Conditional on class membership, knowing one item response gives no additional information about another item response
  - Any association among observed variables is explained entirely by the latent class

## Finite Mixture Models

- LCA models are special cases of **Finite Mixture Models (FMM)**
- A finite mixture model expresses the distribution of **X** as a weighted sum of class-specific distributions:

$$f(\textbf{X}) = \sum_{g=1}^G \eta_g f(\textbf{X}|g)$$

- $\eta_g$ is the mixing proportion for class $g$ (must sum to 1)
- $f(\textbf{X}|g)$ is the class-specific distribution of **X**
- In LCA, $f(\textbf{X}|g)$ is a product of Bernoulli distributions (due to local independence)

## Latent Class Analysis as a FMM

A latent class model for $J$ binary indicators ($j = 1,\ldots,J$) with $C$ classes ($c = 1,\ldots,C$):

$$f(\mathbf{x}_i) = \displaystyle {\sum_{c=1}^{C} \eta_c} \prod_{j=1}^{J} \pi_{jc}^{x_{ij}} \left(1-\pi_{jc}\right)^{1-x_{ij}}$$

- $\eta_c$ — probability that any individual is in class $c$ (must sum to 1)
- $x_{ij}$ — observed binary response of individual $i$ to item $j$
- $\pi_{jc}$ — probability of a positive response ($x_{ij} = 1$) for class $c$ on item $j$

## Estimation Process

- Successfully applying LCA involves answering two key questions:
  1. **How many classes are present?**
     - Fit LCA models with differing numbers of classes
     - Choose based on fit statistics (AIC, BIC, chi-square, entropy)
  2. **What does each class represent?**
     - Inspect the item response probabilities ($\hat{\pi}_{jc}$) for the best-fitting solution
     - Look for meaningful patterns distinguishing the classes

## LCA in R: The `poLCA` Package

- In R, the **`poLCA`** package (Linzer & Lewis, 2011) fits LCA models
  - "**P**olytomous **V**ariable **L**atent **C**lass **A**nalysis"
  - Handles binary and polytomous (multi-category) indicators
  - Uses EM algorithm with multiple random starts
- Install: `install.packages("poLCA")`

## LCA in R: The `poLCA` Package


| Argument | Description |
|----------|-------------|
| `formula` | `cbind(item1, item2, ...) ~ 1` |
| `data` | Data frame; **item values must start at 1**, not 0 |
| `nclass` | Number of latent classes to estimate |
| `nrep` | Number of random starting points (≥ 10 recommended) |
| `verbose` | `FALSE` to suppress iteration output |

# LCA Example

## {.center}

::: {style="text-align: center; font-size: 2em;"}
LCA Example #1
:::

## LCA Example: Macready & Dayton (1977)

::: {.columns}
::: {.column width="60%"}
- Data discussed in Bartholomew and Knott (*Latent Variable Models and Factor Analysis*)
- A four-item binary math test administered to $N = 142$ students
- Macready and Dayton's goal: classify students into two latent groups:
  - **Masters** — students who have mastered the content
  - **Non-masters** — students who have not mastered the content
- We will fit a **2-class LCA model** and interpret the results
:::
::: {.column width="40%"}

| Variable | Description |
|----------|-------------|
| `u1` | Math item 1 (0/1) |
| `u2` | Math item 2 (0/1) |
| `u3` | Math item 3 (0/1) |
| `u4` | Math item 4 (0/1) |


:::
:::

## LCA in R: Data Preparation

```{r}
library(poLCA)

# Load the data (comma-separated, no header)
md_data = read.csv("mddata.dat", header = FALSE)
colnames(md_data) = c("u1", "u2", "u3", "u4")

# poLCA requires values starting at 1 (not 0)
# Recode: 0 -> 1 (incorrect) and 1 -> 2 (correct)
md_data = md_data + 1

# Verify recoding
head(md_data)
table(md_data$u1)
```

::: {.callout-important}
`poLCA` requires each item's values to be **positive integers starting at 1**. For binary items originally coded 0/1, add 1 to recode them as 1/2 before fitting.
:::

## LCA in R: Fitting the 2-Class Model

```{r}
# Define the model formula: all items as indicators, no covariates
lca_formula = cbind(u1, u2, u3, u4) ~ 1

# Fit 2-class LCA with 10 random starting points
set.seed(123)
lca_2class = poLCA(lca_formula,
                   data    = md_data,
                   nclass  = 2,
                   nrep    = 10,
                   verbose = FALSE)

# Print a summary of results
print(lca_2class)
```

- `nrep = 10` runs the algorithm 10 times with different random starts
- The solution with the **highest log-likelihood** is retained
- Increasing `nrep` reduces the chance of stopping at a local maximum

## LCA Parameter Information Types

Three types of information from an LCA model:

1. **Class size estimates** ($\eta_c$)
   - Proportion of the population in each class

2. **Item response probabilities** ($\pi_{jc}$)
   - Probability of a correct response to item $j$ for a person in class $c$

3. **Posterior class probabilities** ($\alpha_{ic}$)
   - Probability that individual $i$ belongs to class $c$, given their responses

## Class Size Estimates ($\eta_c$)

**From `poLCA` output:**

```
Estimated class population shares
 0.5866 0.4134

Predicted class memberships (modal posterior prob.)
 0.5423 0.4577
```

- **Class 1:** approximately **58.7%** of the population ($\eta_1 = 0.587$)
- **Class 2:** approximately **41.3%** of the population ($\eta_2 = 0.413$)

::: {.callout-note}
"Population shares" come from the model. "Predicted memberships" assign each person to their most probable class — these will differ slightly from the model-estimated shares.
:::

## Item Response Probability Estimates ($\pi_{jc}$)

**From `poLCA` output** (Pr(2) = probability of correct response):

```
Conditional item response probabilities by class

$u1      Pr(1)  Pr(2)         $u2      Pr(1)  Pr(2)
 Class 1: 0.247  0.753          Class 1: 0.220  0.780
 Class 2: 0.791  0.209          Class 2: 0.932  0.068

$u3      Pr(1)  Pr(2)         $u4      Pr(1)  Pr(2)
 Class 1: 0.568  0.432          Class 1: 0.292  0.708
 Class 2: 0.982  0.018          Class 2: 0.948  0.052
```

- `Pr(2)` is the probability of a correct response (original code = 1)
- Class 1 has **high** correct-response probabilities; Class 2 has **very low** probabilities

## Interpreting the Classes

| Item | Class 1 $\pi_{j1}$ | Class 2 $\pi_{j2}$ |
|:----:|:---:|:---:|
| u1 | 0.753 | 0.209 |
| u2 | 0.780 | 0.068 |
| u3 | 0.432 | 0.018 |
| u4 | 0.708 | 0.052 |

- **Class 1** — high probability of correct response across all items
  - $\Rightarrow$ **Masters**: students who have mastered the material
- **Class 2** — very low probability of correct response across all items
  - $\Rightarrow$ **Non-masters**: students who have not mastered the material

## Visualizing Item Response Profiles

```{r, echo=FALSE, include=TRUE, eval=TRUE}
library(ggplot2)

# Extract Pr(correct) = Pr(2) for each item and class
probs = data.frame(
  item  = rep(c("u1", "u2", "u3", "u4"), 2),
  class = rep(c("Class 1 (Masters)", "Class 2 (Non-Masters)"), each = 4),
  prob  = c(lca_2class$probs$u1[1, 2],
            lca_2class$probs$u2[1, 2],
            lca_2class$probs$u3[1, 2],
            lca_2class$probs$u4[1, 2],
            lca_2class$probs$u1[2, 2],
            lca_2class$probs$u2[2, 2],
            lca_2class$probs$u3[2, 2],
            lca_2class$probs$u4[2, 2])
)

ggplot(probs, aes(x = item, y = prob, group = class, color = class)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(limits = c(0, 1), name = "P(Correct Response)") +
  labs(x = "Item", title = "LCA Item Response Profiles", color = "Class") +
  theme_bw(base_size = 14)
```

# Assessing Model Fit

## {.center}

::: {style="text-align: center; font-size: 2em;"}
Assessing Model Fit
:::

## Approaches to Assessing LCA Fit

- No single definitive approach — use multiple measures together:
  - **Absolute fit**: model-based $\chi^2$ tests
  - **Relative fit**: information criteria (AIC, BIC)
  - **Classification quality**: entropy
- The goal is to identify a parsimonious, substantively meaningful solution

## Model Chi-Squared Test

- The $\chi^2$ test compares **observed** response pattern frequencies to **model-expected** frequencies
- For $J$ binary items, there are $p = 2^J$ possible response patterns
- Degrees of freedom = number of response patterns $-$ model parameters $-$ 1
- The Pearson statistic:

$$\chi^2_p = \sum_r \frac{(O_r - E_r)^2}{E_r}$$

## Model Chi-Squared Test Limitations

- For $J$ binary items, there are $2^J$ possible response patterns

- **Limitation:** invalid when many cells have small expected frequencies
  - With $J = 4$ items: 16 patterns — feasible
  - With $J = 20$ items: over 1 million patterns — infeasible

## Chi-Squared in R

```{r}
# poLCA automatically computes chi-square tests in the model object
lca_2class$Chisq   # Pearson chi-square statistic
lca_2class$Gsq     # Likelihood ratio G^2 statistic
lca_2class$df      # Degrees of freedom

# Compute p-values
pchisq(lca_2class$Chisq, df = lca_2class$resid.df, lower.tail = FALSE)
pchisq(lca_2class$Gsq,   df = lca_2class$resid.df, lower.tail = FALSE)
```

**Results for the Macready & Dayton 2-class solution:**

```
Pearson chi-square (df = 6):  9.459   p = 0.149
G-squared       (df = 6):     8.966   p = 0.176
```

Neither test is significant — the 2-class model fits the data adequately.

## Log Likelihood

- The **log-likelihood** measures how well the model reproduces the observed data:

$$\log L = \sum_{k=1}^N \log \left( {\sum_{c=1}^{C} \eta_c} \prod_{j=1}^{J} \pi_{jc}^{x_{kj}} \left(1-\pi_{jc}\right)^{1-x_{kj}} \right)$$

## Information Criteria

- The **AIC** and **BIC** penalize the log-likelihood for model complexity:

$$AIC = 2q - 2\log L \qquad BIC = q\log(N) - 2\log L$$

- $q$ = number of free parameters; $N$ = sample size
- **Lower values indicate better fit** (for comparing models with the same data)

## Comparing Models in R

```{r}
# Fit 1-, 2-, and 3-class models
set.seed(123)
lca_1class = poLCA(lca_formula, data = md_data, nclass = 1,
                   nrep = 5,  verbose = FALSE)
lca_2class = poLCA(lca_formula, data = md_data, nclass = 2,
                   nrep = 10, verbose = FALSE)
lca_3class = poLCA(lca_formula, data = md_data, nclass = 3,
                   nrep = 10, verbose = FALSE)

# Compile fit statistics into a table
fit_table = data.frame(
  Classes    = 1:3,
  Parameters = c(lca_1class$npar, lca_2class$npar, lca_3class$npar),
  LogL       = c(lca_1class$llik, lca_2class$llik, lca_3class$llik),
  AIC        = c(lca_1class$aic,  lca_2class$aic,  lca_3class$aic),
  BIC        = c(lca_1class$bic,  lca_2class$bic,  lca_3class$bic)
)
print(fit_table, digits = 2)
```

## Information Criteria: Model Comparison

**Fit statistics for the Macready & Dayton data:**

| Classes | Parameters | Log L | AIC | BIC |
|:---:|:---:|:---:|:---:|:---:|
| 1 | 4 | −373.04 | 754.08 | 766.16 |
| **2** | **9** | **−331.76** | **681.53** | **708.13** |
| 3 | 14 | −331.49 | 690.97 | 733.10 |

- Both AIC and BIC are **lowest for the 2-class solution** ✓
- Adding a third class provides minimal improvement in log-likelihood
- **The 2-class model is preferred** — consistent with Macready and Dayton's hypothesis

## Entropy

- **Entropy** measures how clearly each observation is classified into a class
- Based on posterior class probabilities ($\hat{\alpha}_{ic}$)
- The **relative entropy** (scaled to $[0, 1]$):
  - **E near 1**: high classification certainty (each person clearly belongs to one class)
  - **E near 0**: high uncertainty (posterior probabilities are diffuse across classes)

$$E = 1 - \frac{-\displaystyle\sum_{i=1}^N \sum_{c=1}^C \hat{\alpha}_{ic} \log \hat{\alpha}_{ic}}{N \log C}$$


## Computing Entropy in R

```{r}
# Relative entropy function for poLCA output
compute_entropy = function(posterior) {
  N = nrow(posterior)
  C = ncol(posterior)
  posterior = pmax(posterior, 1e-10)   # avoid log(0)
  entropy_raw = -sum(posterior * log(posterior))
  relative_entropy = 1 - entropy_raw / (N * log(C))
  return(relative_entropy)
}

# Apply to the 2-class model
entropy_2class = compute_entropy(lca_2class$posterior)
cat("Relative Entropy (2-class):", round(entropy_2class, 3))
```


- Entropy of 0.754 indicates **good classification quality**
- Most students are clearly classified as masters or non-masters

# Latent Profile Analysis

## {.center}

::: {style="text-align: center; font-size: 2em;"}
Latent Profile Analysis
:::

## LPA Introduction

- Latent Profile Analysis (LPA) is the **continuous-variable analog to LCA**
  - Also attributed to Lazarsfeld and Henry (1968)
- Uses a set of **continuous variables** rather than binary indicators
- The estimation process is the same: EM algorithm with multiple random starts
- The key difference: **normal (Gaussian) distributions** replace Bernoulli distributions within each class

## LPA Process

- For a specified number of classes, LPA estimates:
  1. **For each class:** the mean ($\mu_{jc}$) and variance ($\sigma^2_{jc}$) for each variable
  2. **For each observation:** the probability of belonging to each class $\alpha_{ic}$
     - These sum to 1 across classes for each person
  3. **Across observations:** the overall class proportions $\eta_c$

- The same three types of information as LCA:
  - Class sizes, class-specific parameters, and posterior probabilities

## LPA Distributional Assumptions

- Within each latent class, the variables are assumed to:
  - Be **independent** (local independence assumption, same as LCA)
  - Follow a **normal distribution** marginally:

$$f(x_i) = \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(\frac{-(x_i - \mu)^2}{2\sigma^2}\right)$$

## LPA Distributional Assumptions

- Different classes have different means — and possibly different variances
- The **variance structure** across classes is an important modeling choice:
  - **Equal variances** across classes (more parsimonious, common default)
  - **Unequal variances** (more flexible, more parameters to estimate)

## Multivariate Normal Distribution

- Because LPA assumes independent normal variables within each class, the joint within-class distribution is **multivariate normal (MVN)** with a *diagonal* covariance matrix

$$f(\textbf{x}) = \frac{1}{(2\pi)^{p/2}|\boldsymbol{\Sigma}|^{1/2}} \exp\left(-\frac{(\textbf{x}-\boldsymbol{\mu})^T\boldsymbol{\Sigma}^{-1}(\textbf{x}-\boldsymbol{\mu})}{2}\right)$$

- $\boldsymbol{\mu}$ = class mean vector
- $\boldsymbol{\Sigma}$ = covariance matrix (**diagonal** for standard LPA — no within-class correlations)
- Standard notation: $\mathbf{x} \sim N_p(\boldsymbol{\mu}, \boldsymbol{\Sigma})$

## Within-Class Covariance Structure

- In standard LPA, within-class covariance is **zero** (local independence):
  - Knowing one variable gives no additional information about another, given class

- When within-class covariance is **non-zero**:
  - Contours become **tilted ellipses**

- The **local independence assumption** can be checked:
  - Fit a model that allows within-class covariances and compare fit via AIC/BIC
  - Persistent residual correlations within classes suggest the assumption is violated

::: {.callout-tip}
If you suspect within-class correlations, try `variances = "equal", covariances = "equal"` (Model 2) and compare fit statistics to Model 1.
:::

## LPA as a Finite Mixture Model

A latent profile model for $J$ continuous variables $(j = 1,\ldots,J)$ with $C$ classes ($c = 1,\ldots,C$):

$$f(\mathbf{x}_i) = \displaystyle {\sum_{c=1}^{C} \eta_c} \prod_{j=1}^{J} \frac{1}{\sqrt{2\pi\sigma^2_{jc}}} \exp\left(\frac{-(x_{ij} - \mu_{jc})^2}{2\sigma^2_{jc}}\right)$$

- $\eta_c$ — class proportion (must sum to 1)
- $\mu_{jc}$ — mean of variable $j$ in class $c$
- $\sigma^2_{jc}$ — variance of variable $j$ in class $c$

## LPA in R: The `tidyLPA` Package

- In R, the **`tidyLPA`** package (Rosenberg et al., 2018) fits LPA models
  - Wraps the `mclust` package with a cleaner interface
  - Install: `install.packages("tidyLPA")`


## LPA in R: The `tidyLPA` Package

| Function | Description |
|----------|-------------|
| `estimate_profiles()` | Fits one or more LPA models |
| `get_fit()` | Extracts fit statistics (log L, AIC, BIC, entropy) |
| `get_estimates()` | Extracts class-specific means and variances |
| `get_data()` | Returns data with posterior probabilities and class assignments |
| `plot_profiles()` | Creates a class profile plot |

## LPA Variance Model Options


| Model | Variances | Covariances |
|:---:|:---:|:---:|
| 1 (default) | Equal across classes | Zero (diagonal) |
| 2 | Equal across classes | Equal across classes |
| 3 | **Varying** across classes | Zero (diagonal) |

- **Model 1** matches the standard LPA assumption: equal variances, no within-class covariances

## LPA Variance Model Options


| Model | Variances | Covariances |
|:---:|:---:|:---:|
| 4 | **Varying** across classes | **Varying** across classes |
| 5 | Equal across classes | **Varying** across classes |
| 6 | **Varying** across classes | Equal across classes |


- More complex models add parameters; use information criteria to compare

# LPA Example

## {.center}

::: {style="text-align: center; font-size: 2em;"}
LPA Example: Fisher's Iris Data
:::

## LPA Example: Fisher's Iris Data

- Fisher (1936) introduced a dataset of measurements on 150 iris flowers
- Four continuous measurements per flower:

| Variable | Measurement |
|----------|-------------|
| x1 | Sepal length (cm) |
| x2 | Sepal width (cm) |
| x3 | Petal length (cm) |
| x4 | Petal width (cm) |

## LPA Example: Fisher's Iris Data

- Three species: *Iris setosa*, *I. versicolor*, *I. virginica* (50 flowers each; $N = 150$)
- We treat species membership as **unknown** and attempt to recover it via LPA

## LPA in R: Loading Data and Fitting Models

```{r, echo=FALSE}
library(tidyLPA)

# Load the Iris data (space-separated, no header, no species label)
iris_dat = read.table("iris.dat", header = FALSE)
colnames(iris_dat) = c("x1", "x2", "x3", "x4")

# Fit LPA models with 2, 3, and 4 classes (Model 1: equal variances)
lpa_models = estimate_profiles(iris_dat,
                               n_profiles  = 2:4,
                               variances   = "equal",
                               covariances = "zero")

# Compare fit statistics across models
get_fit(lpa_models)
```

::: {.callout-note}
`variances = "equal"` and `covariances = "zero"` specify Model 1 — equal variances across classes, no within-class covariances. This is the standard LPA assumption.
:::

## Model Comparison

**Fit statistics for the Iris LPA (Model 1):**

| Classes | Parameters | Log L | AIC | BIC | Entropy |
|:---:|:---:|:---:|:---:|:---:|:---:|
| 2 | 13 | −488.92 | 1003.83 | 1042.97 | 0.991 |
| **3** | **18** | **−361.43** | **758.85** | **813.04** | **0.957** |
| 4 | 23 | −310.12 | 666.23 | 735.48 | 0.945 |

- AIC and BIC both favor the **4-class solution**
- But the data were generated from **exactly 3 species**
- This illustrates a known problem with information criteria in mixture modeling:
  - They can favor over-extraction (more classes than truly exist)
  - **Substantive knowledge and external validation are essential**
- We select the **3-class solution** based on prior knowledge

## Fitting the 3-Class Model

```{r, echo=FALSE}
# Fit the 3-class LPA model
lpa_3class = estimate_profiles(iris_dat,
                               n_profiles  = 3,
                               variances   = "equal",
                               covariances = "zero")

# View class-specific estimates
get_estimates(lpa_3class)
```

## Fitting the 3-Class Model

```{r, echo=FALSE}
# Add posterior probabilities and class assignments to the data
lpa_3class_data = get_data(lpa_3class)
head(lpa_3class_data)
```

## The 3-Class Solution: Class Sizes

**From `get_fit()` output:**

```
FINAL CLASS COUNTS AND PROPORTIONS
   Class 1:  50.000  (0.333)
   Class 2:  54.888  (0.366)
   Class 3:  45.112  (0.301)
```

- Three approximately equal classes, consistent with the true species distribution (50 per species)
- The model has correctly identified the rough size structure

## The 3-Class Solution: Class Means

**Class-specific means from `get_estimates()`:**

| Variable | Class 1 | Class 2 | Class 3 |
|:---------|:---:|:---:|:---:|
| x1 (Sepal Length) | 5.01 | 5.92 | 6.68 |
| x2 (Sepal Width) | 3.43 | 2.75 | 3.02 |
| x3 (Petal Length) | 1.46 | 4.33 | 5.61 |
| x4 (Petal Width) | 0.25 | 1.35 | 2.07 |

- **Class 1**: Very small petals, wide sepals → *Iris setosa*
- **Class 2**: Medium-sized flowers, narrower sepals → *Iris versicolor*
- **Class 3**: Largest petals and sepals → *Iris virginica*

## Visualizing Class Profiles

```{r}
# Built-in profile plot (quickest option)
plot_profiles(lpa_3class, add_line = TRUE, rawdata = FALSE)
```
## Visualizing Class Profiles

```{r, echo=FALSE}
# Custom ggplot2 profile plot (more control)
library(ggplot2)

# get_estimates() returns: Model, Classes, Class, Parameter, Estimate, SE, p, Item
ests = get_estimates(lpa_3class)
ests_means = ests[ests$Category == "Means", ]

# Label classes by species
ests_means$Class_label = factor(ests_means$Class,
                                levels = 1:3,
                                labels = c("Class 1 (setosa)",
                                           "Class 2 (versicolor)",
                                           "Class 3 (virginica)"))

ggplot(ests_means, aes(x = Parameter, y = Estimate,
                       group = Class_label, color = Class_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(x = "Variable", y = "Mean",
       title = "LPA Profile Plot: 3-Class Solution",
       color = "Class") +
  theme_bw(base_size = 14)
```

## Evaluating Classification Quality

```{r, eval=FALSE}
# Average posterior probability for most likely class assignment
# (a measure of classification certainty)
lpa_3class_data = get_data(lpa_3class)

avg_posterior = tapply(lpa_3class_data$Probability,
                       lpa_3class_data$Class,
                       mean)
print(round(avg_posterior, 3))
```

```
Class 1  Class 2  Class 3
  1.000    0.970    0.966
```

- Class 1 (setosa) is perfectly separated from the others
- Classes 2 and 3 have 3–4% overlap — the two larger species are harder to distinguish
- Matches the known botanical overlap between *versicolor* and *virginica*

# Concluding Remarks

## {.center}

::: {style="text-align: center; font-size: 2em;"}
Concluding Remarks
:::

## LCA vs. LPA: Comparison

| Feature | LCA | LPA |
|:--------|:----|:----|
| Indicator type | Binary / categorical | Continuous |
| Within-class distribution | Bernoulli | Normal |
| Class parameters | Item probabilities ($\pi_{jc}$) | Means ($\mu_{jc}$) and variances ($\sigma^2_{jc}$) |
| R package | `poLCA` | `tidyLPA` |

## Extensions of These Methods

- Many extensions exist in recent empirical research:
  - **Growth Mixture Models** — detect groups with differing growth trajectories over time
  - **Diagnostic Classification Models** — confirmatory LCA specifying class characteristics *prior* to analysis; used in psychological and educational testing
  - **Mixture IRT Models** — combine item response theory with latent class structure
  - **Mixture Regression Models** — latent classes defined by distinct regression relationships
  - **General Finite Mixture Models** — virtually any statistical distribution can form a mixture model

## Concluding Remarks

- LCA and LPA are **model-based techniques** for identifying latent clusters
- Both methods make explicit, testable assumptions about the data
- Key practical challenges:
  - The number of classes must be inferred — no single definitive criterion
  - Information criteria often favor **over-extraction** (too many classes)
  - Multiple local maxima are possible — always use many random starts (`nrep`)
- **Use these methods carefully** — they are exploratory and prone to spurious findings
  - Validate results with substantive knowledge
  - Seek convergence across multiple fit indices
  - Replicate with independent samples when possible

## References

- Lazarsfeld, P. F. & Henry, N. W. (1968). *Latent structure analysis*. Houghton Mifflin.
- Linzer, D. A. & Lewis, J. B. (2011). poLCA: An R package for polytomous variable latent class analysis. *Journal of Statistical Software, 42*(10), 1–29.
- Macready, G. B. & Dayton, C. M. (1977). The use of probabilistic models in the assessment of mastery. *Journal of Educational Statistics, 2*(2), 99–120.
- Rosenberg, J. M., van Lissa, C. J., Beymer, P. N., Anderson, D. J., Schell, M. J., & Schmidt, J. A. (2018). tidyLPA: An R package to easily carry out latent profile analysis (LPA). *Journal of Open Source Software, 3*(30), 978.
