---
title: "Missing Data Methods"
subtitle: "A Brief Introduction"
author: "Jonathan Templin"
format:
  revealjs:
    theme: simple
    css: clemson.css
    slide-number: true
    logo: clemson_coe.png
    transition: slide
    chalkboard: true
    preview-links: auto
execute:
  echo: false
  eval: true
---

## About This Lecture {.center}

::: {.callout-note}
**This lecture is a broad overview only.**
Missing data is a deep field with a rich methodological literature.
Our goal today is to build intuition, introduce vocabulary, and demonstrate
two modern workflows — not to provide exhaustive technical coverage.
:::

> *"The problem of missing data is one of the most pervasive in the practice of statistics."*
> — Little & Rubin (2002)

::: notes
Set expectations up front. PhD students are often frustrated when a lecture doesn't go deep enough — acknowledge this explicitly. Point them toward Little & Rubin (2002) and van Buuren's free online book *Flexible Imputation of Missing Data* for deeper treatment. Today is about making sure they can recognize missingness mechanisms, critique naive approaches, and implement two principled methods.
:::

---

## Roadmap

::: {.incremental}
1. **Why missing data matters** — it's not just inconvenience
2. **Taxonomy** — MCAR, MAR, MNAR
3. **The old (wrong) ways** — what not to do and why
4. **Modern Method 1:** Full Information Maximum Likelihood (FIML) for path models
5. **Modern Method 2:** Multiple Imputation by Chained Equations (MICE)
6. **Practical guidance** and wrap-up
:::

::: notes
Walk through the roadmap briefly. Emphasize that sections 3–5 build on each other — students need to understand why the naive approaches fail before they can appreciate why FIML and MICE are better. The path analysis example will be fully simulated so the ground truth is known, which makes the comparison clean.
:::

---

# Why Missing Data Matters


---

## Missing Data Is Everywhere

Real data is almost never complete. Common sources:

- Survey non-response (certain groups less likely to respond)
- Attrition in longitudinal studies
- Instrument or sensor failure
- Routing in adaptive questionnaires
- Merging data from multiple sources


**The danger:** Most software defaults to complete-case analysis — silently dropping rows with any missing values.


## Two Consequences of Mishandling Missing Data

- **Bias** — estimates that are systematically wrong
- **Inefficiency** — throwing away information, inflating standard errors


---

## A Simple Motivating Example

```{r}
#| label: motivating-example
#| echo: false
#| fig-width: 9
#| fig-height: 4.5

set.seed(2024)
n = 400
x = rnorm(n, mean = 50, sd = 10)
y = 2 + 0.6 * x + rnorm(n, sd = 8)
df_full = data.frame(x = x, y = y)

# Introduce MNAR: high-y values are more likely to be missing
miss_prob = plogis((y - mean(y)) / sd(y))
df_mnar = df_full
df_mnar$y[runif(n) < miss_prob * 0.7] = NA

par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))

plot(df_full$x, df_full$y,
     pch = 16, col = "#2980b9", cex = 0.7,
     xlab = "X", ylab = "Y", main = "Complete Data")
abline(lm(y ~ x, data = df_full), col = "#e74c3c", lwd = 2)

obs_idx = !is.na(df_mnar$y)
plot(df_mnar$x[obs_idx], df_mnar$y[obs_idx],
     pch = 16, col = "#2980b9", cex = 0.7,
     xlab = "X", ylab = "Y", main = "After Ignoring Missing (MNAR)")
abline(lm(y ~ x, data = df_mnar), col = "#e74c3c", lwd = 2)
abline(lm(y ~ x, data = df_full), col = "#7f8c8d", lwd = 2, lty = 2)
legend("topleft", legend = c("Observed fit", "True fit"),
       col = c("#e74c3c", "#7f8c8d"), lty = c(1, 2), lwd = 2, cex = 0.8)
```

Even a moderate amount of non-random missingness visibly biases our estimate.


---

# Taxonomy of Missing Data

---

## The Three Mechanisms (Rubin, 1976)

Let $Y$ be the full data matrix and $R$ be the **missingness indicator** (1 = observed, 0 = missing).


| Mechanism | Definition | Intuition |
|-----------|-----------|-----------|
| **MCAR** | $P(R \mid Y) = P(R)$ | Missingness is random coin flip, unrelated to any data |

## The Three Mechanisms (Rubin, 1976)

| Mechanism | Definition | Intuition |
|-----------|-----------|-----------|
| **MAR** | $P(R \mid Y) = P(R \mid Y_{obs})$ | Missingness depends on *observed* data, not on the missing values themselves |

## The Three Mechanisms (Rubin, 1976)

| Mechanism | Definition | Intuition |
|-----------|-----------|-----------|
| **MNAR** | $P(R \mid Y) = P(R \mid Y_{obs}, Y_{mis})$ | Missingness depends on the *missing values themselves* |

::: {.callout-important}
MCAR ⊂ MAR. MCAR is a special case of MAR. MNAR is the hardest — and most realistic — case.
:::

---

## MCAR — Missing Completely at Random

**Definition:** The probability of missingness is the same for all units, regardless of any data values.

```{r}
#| label: mcar-demo
set.seed(42)
n = 300
x = rnorm(n)
y = 1.5 * x + rnorm(n)
df = data.frame(x = x, y = y)

# MCAR: randomly remove 30% of y values, unrelated to anything
df_mcar = df
df_mcar$y[sample(1:n, size = round(0.3 * n))] = NA

cat("Proportion missing:", mean(is.na(df_mcar$y)), "\n")
cat("Mean of x for observed y:  ", round(mean(df_mcar$x[!is.na(df_mcar$y)]), 3), "\n")
cat("Mean of x for missing y:   ", round(mean(df_mcar$x[is.na(df_mcar$y)]), 3), "\n")
```

. . .

Under MCAR, observed and missing groups look similar on all covariates.

**Test:** Little's MCAR test (`naniar::mcar_test()` in R) — though low power in small samples.

::: notes
The two means of X should be nearly identical — that's the hallmark of MCAR. In practice, true MCAR is rare. It would require, for example, that a survey question was accidentally omitted on a random subset of forms. More commonly, missingness has some systematic component (MAR or MNAR).
:::

---

## MAR — Missing at Random

**Definition:** Missingness depends on *observed* variables, but NOT on the variable that is missing.

```{r}
#| label: mar-demo
set.seed(42)
n = 300
x = rnorm(n)
y = 1.5 * x + rnorm(n)
df = data.frame(x = x, y = y)

# MAR: missingness in y depends on x (observed), not on y itself
df_mar = df
miss_prob_mar = plogis(1.5 * df$x)   # higher x → more likely y is missing
df_mar$y[runif(n) < miss_prob_mar] = NA

cat("Proportion missing:", round(mean(is.na(df_mar$y)), 3), "\n")
cat("Mean of x for observed y:  ", round(mean(df_mar$x[!is.na(df_mar$y)]), 3), "\n")
cat("Mean of x for missing y:   ", round(mean(df_mar$x[is.na(df_mar$y)]), 3), "\n")
```

. . .

X predicts missingness — but once we condition on X, missingness is random.

**Key implication:** Methods that use the full covariate structure (FIML, MICE) can recover unbiased estimates under MAR.


---

## MNAR — Missing Not at Random

**Definition:** Missingness depends on the *unobserved values themselves*, even after conditioning on all observed data.

```{r}
#| label: mnar-demo
set.seed(42)
n = 300
x = rnorm(n)
y = 1.5 * x + rnorm(n)
df = data.frame(x = x, y = y)

# MNAR: missingness in y depends on y itself
df_mnar = df
miss_prob_mnar = plogis(1.5 * df$y)   # higher y → more likely y is missing
df_mnar$y[runif(n) < miss_prob_mnar] = NA

cat("Proportion missing:", round(mean(is.na(df_mnar$y)), 3), "\n")
cat("Mean of OBSERVED y:  ", round(mean(df_mnar$y, na.rm = TRUE), 3), "\n")
cat("Mean of TRUE y:      ", round(mean(df$y), 3), "\n")
```

. . .

::: {.callout-warning}
MNAR is the hardest case. Neither FIML nor MICE provides a general solution.
Addressing MNAR requires *sensitivity analyses* and *explicit models for the missingness mechanism*.
:::


---

## Summary: The Three Mechanisms

```{r}
#| label: mechanism-summary
#| echo: false
#| fig-width: 10
#| fig-height: 4

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
set.seed(42)
n = 200
x_vals = rnorm(n)
y_vals = 1.5 * x_vals + rnorm(n)

# MCAR
obs_mcar = sample(c(TRUE, FALSE), n, replace = TRUE, prob = c(0.7, 0.3))
plot(x_vals, y_vals, pch = ifelse(obs_mcar, 16, 4),
     col = ifelse(obs_mcar, "#2980b9", "#e74c3c"),
     main = "MCAR", xlab = "X", ylab = "Y", cex = 0.8)
legend("topleft", pch = c(16, 4), col = c("#2980b9", "#e74c3c"),
       legend = c("Observed", "Missing"), cex = 0.75)

# MAR
miss_mar = runif(n) < plogis(1.5 * x_vals)
obs_mar = !miss_mar
plot(x_vals, y_vals, pch = ifelse(obs_mar, 16, 4),
     col = ifelse(obs_mar, "#2980b9", "#e74c3c"),
     main = "MAR (missing when X is high)", xlab = "X", ylab = "Y", cex = 0.8)

# MNAR
miss_mnar = runif(n) < plogis(1.5 * y_vals)
obs_mnar = !miss_mnar
plot(x_vals, y_vals, pch = ifelse(obs_mnar, 16, 4),
     col = ifelse(obs_mnar, "#2980b9", "#e74c3c"),
     main = "MNAR (missing when Y is high)", xlab = "X", ylab = "Y", cex = 0.8)
```

::: notes
This visual comparison is worth spending a minute on. MCAR: missing values scattered uniformly. MAR: missing values cluster on the right (high X) but spread across Y values. MNAR: missing values cluster at the top (high Y), which is exactly what we cannot account for without modeling the missingness directly.
:::

---

# The Early (Wrong) Ways


---

## Approach 1: Listwise (Complete-Case) Deletion

**What it does:** Drop any observation with a missing value on *any* variable in the model.

```{r}
#| label: listwise
set.seed(42)
df_example = data.frame(
  id  = 1:10,
  x1  = c(2.1, 3.5, NA, 1.8, 4.2, 2.9, NA, 3.1, 2.7, 4.0),
  x2  = c(1.0, NA, 2.2, 1.5, NA, 2.8, 1.9, NA, 2.3, 1.7),
  y   = c(5.1, 6.8, 4.9, 5.5, 7.2, 6.1, 5.8, 6.4, 5.7, 7.0)
)

complete_cases = df_example[complete.cases(df_example), ]
cat("Original N:", nrow(df_example), "\n")
cat("After listwise deletion N:", nrow(complete_cases), "\n")
cat("Percent retained:", round(100 * nrow(complete_cases) / nrow(df_example), 1), "%\n")
```

. . .

**Problems:**

- Requires MCAR for unbiased estimates — rarely justified
- Loses substantial power; in wide datasets with many variables, near-total loss of data is possible
- Dropping rows changes the sample — who is left?

---

## Approach 2: Pairwise Deletion

**What it does:** Use all available observations for each pairwise calculation (e.g., each cell of a correlation matrix).

```{r}
#| label: pairwise
set.seed(42)
n = 100
x1 = rnorm(n)
x2 = 0.5 * x1 + rnorm(n)
x3 = 0.4 * x2 + rnorm(n)

df_pw = data.frame(x1 = x1, x2 = x2, x3 = x3)
df_pw$x2[sample(1:n, 20)] = NA
df_pw$x3[sample(1:n, 25)] = NA

# Pairwise correlation (default in cor() with use = "pairwise.complete.obs")
cor_pairwise = cor(df_pw, use = "pairwise.complete.obs")
round(cor_pairwise, 3)
```

. . .

**Problem:** Each correlation uses a different subset of the data.

- The resulting correlation matrix may **not be positive semi-definite** — it can be mathematically impossible.
 - SEM/path models will fail or produce nonsensical results.


---

## Approach 3: Mean Imputation

**What it does:** Replace each missing value with the variable's observed mean.

```{r}
#| label: mean-imputation
set.seed(42)
n = 200
y_true = rnorm(n, mean = 5, sd = 2)
y_obs = y_true
y_obs[sample(1:n, 60)] = NA   # 30% missing (MCAR for this demo)

y_mean_imp = y_obs
y_mean_imp[is.na(y_obs)] = mean(y_obs, na.rm = TRUE)

cat("True SD:         ", round(sd(y_true), 3), "\n")
cat("Observed SD:     ", round(sd(y_obs, na.rm = TRUE), 3), "\n")
cat("After mean imp SD:", round(sd(y_mean_imp), 3), "\n")
```

. . .

**Problems:**

- **Artificially deflates variance** — attenuates all correlations and regression coefficients
- Distorts the distribution (spike at the mean)
- Standard errors are wrong — the imputed values are treated as real observations

---

## Approach 4: Single (Regression) Imputation

**What it does:** Predict missing values from a regression on observed variables. Slightly better than mean imputation, but still flawed.

```{r}
#| label: single-imputation
set.seed(42)
n = 200
x = rnorm(n)
y = 1.5 * x + rnorm(n, sd = 1.5)
df_si = data.frame(x = x, y = y)
df_si$y[sample(1:n, 60)] = NA

# Fit regression on observed cases
reg_model = lm(y ~ x, data = df_si, na.action = na.omit)
# Impute missing
df_si$y_imp = df_si$y
missing_idx = is.na(df_si$y)
df_si$y_imp[missing_idx] = predict(reg_model,
                                    newdata = df_si[missing_idx, ])

cat("True residual SD:    ", round(sd(y - (1.5 * x)), 3), "\n")
cat("Imputed residual SD: ", round(sd(df_si$y_imp - predict(lm(y_imp ~ x, data = df_si))), 3), "\n")
```

. . .

**Problem:** Single imputation replaces each missing value with a *point estimate* — no residual noise.

- This **understates uncertainty** and inflates correlations (imputed values lie perfectly on the regression line).



---

# Full Information Maximum Likelihood (FIML)

---

## How FIML Works (Conceptually)

Under FIML (also called *direct ML* or *raw ML*):

- Each observation contributes to the likelihood using **only its observed variables**
- No data are deleted or imputed
- The full model is estimated simultaneously with the missingness

## How FIML Works (Conceptually)

**The likelihood for observation $i$ with observed variables $Y_i^{obs}$:**

$$\ell_i(\theta) = -\frac{1}{2} \left[ \log |\Sigma_i| + (Y_i^{obs} - \mu_i)^\top \Sigma_i^{-1} (Y_i^{obs} - \mu_i) \right]$$

where $\mu_i$ and $\sigma_i$ are the model-implied mean and covariance for the *observed subset* of variables for case $i$.

. . .

::: {.callout-tip}
**Key advantage:** FIML uses the covariance structure to "borrow strength" from observed variables to recover information about missing ones — all within the model itself.
:::


---

## Path Analysis Example: Simulating the Data

We simulate a simple three-variable path model with a known structure:

$$X_1 \rightarrow X_2 \rightarrow Y, \quad X_1 \rightarrow Y$$

```{r}
#| label: simulate-data
set.seed(2025)
n = 500

# Exogenous predictor
x1 = rnorm(n, mean = 0, sd = 1)

# Mediator: caused by x1
x2 = 0.50 * x1 + rnorm(n, sd = sqrt(1 - 0.50^2))

# Outcome: caused by both x1 and x2
y  = 0.30 * x1 + 0.55 * x2 + rnorm(n, sd = sqrt(1 - 0.30^2 - 0.55^2 - 2*0.30*0.55*0.50))

df_full = data.frame(x1 = x1, x2 = x2, y = y)

cat("True path coefficients:\n")
cat("  x2 ~ x1 : 0.50\n")
cat("  y  ~ x1 : 0.30\n")
cat("  y  ~ x2 : 0.55\n")
cat("N =", nrow(df_full), "| No missing values\n")
```


---

## Introducing MAR Missingness

We introduce missingness in $Y$ and $X_2$ that depends on observed $X_1$:

```{r}
#| label: introduce-missingness
set.seed(2025)
df_miss = df_full

# y is missing when x1 is high (MAR: depends on observed x1)
prob_miss_y = plogis(1.2 * df_full$x1)
df_miss$y[runif(n) < prob_miss_y * 0.55] = NA

# x2 is missing when x1 is low (MAR: depends on observed x1)
prob_miss_x2 = plogis(-1.0 * df_full$x1)
df_miss$x2[runif(n) < prob_miss_x2 * 0.40] = NA

cat("Missing in y: ", sum(is.na(df_miss$y)),
    "(", round(100 * mean(is.na(df_miss$y)), 1), "%)\n")
cat("Missing in x2:", sum(is.na(df_miss$x2)),
    "(", round(100 * mean(is.na(df_miss$x2)), 1), "%)\n")
cat("Complete cases:", sum(complete.cases(df_miss)),
    "(", round(100 * mean(complete.cases(df_miss)), 1), "%)\n")
```

. . .

About **half the sample** is incomplete. Complete-case analysis would throw away a large fraction of observations.


---

## Fitting the Path Model: Complete Cases vs. FIML

```{r}
#| label: path-model
#| output: false
library(lavaan)

path_model = "
  x2 ~ x1
  y  ~ x1 + x2
"

# Complete-case analysis (lavaan default drops missing rows)
fit_cc = sem(path_model, data = df_miss, missing = "listwise")

# FIML
fit_fiml = sem(path_model, data = df_miss, missing = "fiml")
```

```{r}
#| label: path-results
results_cc   = parameterEstimates(fit_cc,   standardized = FALSE)
results_fiml = parameterEstimates(fit_fiml, standardized = FALSE)

# Extract regression paths only
paths_cc   = results_cc[results_cc$op == "~",
                         c("lhs", "op", "rhs", "est", "se")]
paths_fiml = results_fiml[results_fiml$op == "~",
                           c("lhs", "op", "rhs", "est", "se")]

paths_cc$method   = "Complete Case"
paths_fiml$method = "FIML"

comparison = rbind(paths_cc, paths_fiml)
comparison$true_value = c(0.50, 0.30, 0.55, 0.50, 0.30, 0.55)
comparison$bias = round(comparison$est - comparison$true_value, 4)
comparison[, c("lhs", "rhs", "method", "true_value", "est", "se", "bias")]
```


---

## FIML Results: Visual Comparison

```{r}
#| label: fiml-visual
#| echo: false
#| fig-width: 9
#| fig-height: 5

path_labels = c("x2~x1 (true=0.50)", "y~x1 (true=0.30)", "y~x2 (true=0.55)")
true_vals   = c(0.50, 0.30, 0.55)

cc_est   = paths_cc$est
cc_se    = paths_cc$se
fi_est   = paths_fiml$est
fi_se    = paths_fiml$se

y_pos = c(3, 2, 1)
par(mar = c(5, 10, 3, 2))
plot(NA, xlim = c(-0.1, 1.0), ylim = c(0.5, 3.5),
     yaxt = "n", xlab = "Path Coefficient Estimate", ylab = "",
     main = "Complete Case vs. FIML Path Estimates")

axis(2, at = y_pos, labels = path_labels, las = 1, cex.axis = 0.85)
abline(v = true_vals, col = "#27ae60", lty = 2, lwd = 1.5)

offset = 0.15
for (i in 1:3) {
  # Complete case
  points(cc_est[i], y_pos[i] + offset, pch = 16, col = "#e74c3c", cex = 1.3)
  lines(c(cc_est[i] - 1.96 * cc_se[i], cc_est[i] + 1.96 * cc_se[i]),
        c(y_pos[i] + offset, y_pos[i] + offset), col = "#e74c3c", lwd = 2)

  # FIML
  points(fi_est[i], y_pos[i] - offset, pch = 17, col = "#2980b9", cex = 1.3)
  lines(c(fi_est[i] - 1.96 * fi_se[i], fi_est[i] + 1.96 * fi_se[i]),
        c(y_pos[i] - offset, y_pos[i] - offset), col = "#2980b9", lwd = 2)
}

legend("topright", pch = c(16, 17), col = c("#e74c3c", "#2980b9"),
       legend = c("Complete Case", "FIML"), cex = 0.9)
legend("topleft", lty = 2, col = "#27ae60",
       legend = "True value", cex = 0.9)
```


---

## FIML: Key Practical Notes

::: {.incremental}
- **Requires normality** — technically assumes multivariate normal data; robust versions (`estimator = "MLR"`) handle modest violations
- **Works within the model** — missingness is handled at the model level, no pre-processing needed
- **Assumes MAR** — if data are MNAR, FIML estimates are still biased
- **Best for SEM/path models** — when you're already in the lavaan/SEM framework, FIML is the natural choice
- **No imputed datasets** — results cannot be used outside the model; if you need imputed data for downstream analyses, use MICE
:::



# Multiple Imputation by Chained Equations (MICE)

---

## The MICE Algorithm (Overview)

**Three stages:**

::: {.incremental}
1. **Impute** — Create $m$ complete datasets by drawing from the conditional distribution of each missing variable given all others. Each variable gets its own imputation model.

2. **Analyze** — Fit the model of interest to each of the $m$ imputed datasets separately.

3. **Pool** — Combine the $m$ sets of estimates using **Rubin's Rules** to produce a single set of estimates and standard errors that reflect both within- and between-imputation uncertainty.
:::


## The Key in the MICE Algorithm

**Why $m$ datasets and not one?**

- Single imputation underestimates uncertainty. 
- The variation *between* imputed datasets captures our uncertainty about the missing values.


---

## The Chained Equations (FCS) Approach

MICE imputes each variable in turn, conditioning on all others:

```
Initialize: fill in starting values for all missing data

For each iteration:
  For each variable Xj with missing values:
    1. Fit model: Xj ~ all other variables (observed + currently imputed)
    2. Draw from the posterior predictive distribution
    3. Replace missing values of Xj with the draws

Repeat for many cycles until convergence
Return the imputed values from the final cycle
```

. . .

::: {.callout-tip}
Each variable can have its **own imputation model** (linear regression, logistic regression, PMM, etc.) — this is the "chained" part. This makes MICE very flexible for mixed variable types.
:::

---

## MICE in R: Setup and Imputation

```{r}
#| label: mice-setup
library(mice)

# Use the same MAR dataset from the path analysis section
cat("Variables with missing data:\n")
print(colSums(is.na(df_miss)))

# Run MICE with m = 20 imputed datasets, PMM for continuous variables
imp = mice(df_miss,
           m       = 20,        # number of imputed datasets
           method  = "pmm",     # predictive mean matching
           maxit   = 10,        # iterations per imputation
           seed    = 2025,
           printFlag = FALSE)

summary(imp)
```


---

## Checking MICE Convergence

```{r}
#| label: mice-convergence
#| fig-width: 9
#| fig-height: 4
plot(imp, layout = c(2, 2))
```


---

## MICE: Analyzing the Imputed Datasets

```{r}
#| label: mice-analysis
# Fit the path-like model (as a regression, for simplicity) on each imputed dataset

# Model 1: x2 ~ x1
fit_x2 = with(imp, lm(x2 ~ x1))

# Model 2: y ~ x1 + x2
fit_y  = with(imp, lm(y ~ x1 + x2))

# Pool using Rubin's Rules
pooled_x2 = pool(fit_x2)
pooled_y  = pool(fit_y)

cat("=== Pooled: x2 ~ x1 ===\n")
summary(pooled_x2)[, c("term", "estimate", "std.error", "statistic", "p.value")]
```


---

## MICE: Pooled Results for the Outcome Model

```{r}
#| label: mice-pooled-y
cat("=== Pooled: y ~ x1 + x2 ===\n")
print(summary(pooled_y)[, c("term", "estimate", "std.error", "statistic", "p.value")])

cat("\n=== Fraction of Missing Information (FMI) ===\n")
print(pooled_y$pooled[, c("term", "fmi", "lambda")])
```


---

## Three-Way Comparison: Complete Case vs. FIML vs. MICE

```{r}
#| label: three-way
#| echo: false

# True values
true_b = c("x2~x1" = 0.50, "y~x1" = 0.30, "y~x2" = 0.55)

# Complete case
cc_x2 = coef(lm(x2 ~ x1, data = df_miss[complete.cases(df_miss), ]))
cc_y  = coef(lm(y  ~ x1 + x2, data = df_miss[complete.cases(df_miss), ]))

# MICE pooled
mx2 = summary(pooled_x2)
my  = summary(pooled_y)

results_df = data.frame(
  Path       = c("x2 ~ x1", "y ~ x1", "y ~ x2"),
  True       = c(0.50, 0.30, 0.55),
  CC         = round(c(cc_x2["x1"], cc_y["x1"], cc_y["x2"]), 3),
  FIML       = round(c(paths_fiml$est[1], paths_fiml$est[2], paths_fiml$est[3]), 3),
  MICE       = round(c(mx2$estimate[mx2$term=="x1"],
                       my$estimate[my$term=="x1"],
                       my$estimate[my$term=="x2"]), 3)
)

results_df$Bias_CC   = round(results_df$CC   - results_df$True, 3)
results_df$Bias_FIML = round(results_df$FIML - results_df$True, 3)
results_df$Bias_MICE = round(results_df$MICE - results_df$True, 3)

print(results_df)
```

. . .

Both FIML and MICE substantially reduce bias relative to complete-case analysis — and the estimates from the two modern methods are quite similar to each other.


# Practical Guidance & Wrap-Up

## Which Method Should You Use?

| Situation | Recommendation |
|-----------|----------------|
| Fitting a SEM / path model | **FIML** via `missing = "fiml"` in lavaan |
| Regression, multilevel, GLM — mixed data types | **MICE** (`mice` package) |
| Need to explore imputed data, check distributions | **MICE** (can inspect imputed datasets) |
| MNAR suspected | Sensitivity analysis; selection or pattern-mixture models |
| < 5% missing, MCAR plausible | Listwise may be acceptable — but still check |

## Which Method Should You Use?

::: {.callout-important}
**Never** use mean imputation in primary analyses reported in a peer-reviewed paper.
If a reviewer or collaborator suggests it, cite Little & Rubin (2002) and explain why.
:::

---

## Key Assumptions to Check


1. **Justify MAR** — ask: *conditional on what I observed, is there any remaining reason values would be missing?* More covariates → MAR more plausible.

2. **Check MICE convergence** — trace plots should show no trends across iterations.

3. **Inspect imputed values** — do imputed values look plausible? PMM helps here by constraining imputations to the observed range.

---

## Key Assumptions to Check



4. **Report what you did** — many papers are vague about missing data handling. Report: % missing per variable, assumed mechanism, method used, number of imputed datasets if MICE.

5. **Sensitivity to MNAR** — if results hinge on MAR, conduct at least one sensitivity analysis relaxing this assumption.

---

## Summary

::: {.incremental}
- Missing data has three mechanisms: **MCAR**, **MAR**, and **MNAR** — the mechanism determines which methods are valid.
- Naive methods (listwise, pairwise, mean imputation) are **biased or inefficient** under anything other than MCAR, and should be avoided for primary analyses.
- **FIML** is the preferred approach within SEM/path models — uses all available data without imputation.
- **MICE** is a general-purpose multiple imputation framework — flexible, transparent, and widely applicable.
- Both methods assume **MAR** — sensitivity analyses are needed if MNAR is plausible.
- This was a **broad overview** — see the references for deeper treatment.
:::

